---
title: "Files in, CSV out"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    df_print: kable
vignette: >
  %\VignetteIndexEntry{Files in, CSV out}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(width = 92)
```

> Excel workbooks, delimited files, and what lands in the output.

```{r}
library(jtb2026)
```

Both search functions take their `data` argument straight to
`jtb_read_table()`, so anything that reads works in either. Nothing about the
search changes with the input type — a path is simply a table you have not
loaded yet.

## What `data` accepts

| `data` | What happens |
| --- | --- |
| A `data.frame`, tibble or `data.table` | Used as-is |
| A matrix | Converted column-wise to a data frame |
| `"file.xlsx"`, `.xlsm`, `.xls` | `readxl::read_excel()`, honouring `sheet` |
| `"file.csv"` | `utils::read.csv()` |
| `"file.tsv"`, `"file.txt"` | `utils::read.delim()` |
| Anything else | An error naming the extensions that do work |

The package ships the mock table in both formats, so the examples below are
runnable:

```{r extdata}
xlsx <- system.file("extdata", "site_reports.xlsx", package = "jtb2026")
csv <- system.file("extdata", "site_reports.csv", package = "jtb2026")

basename(c(xlsx, csv))
```

## Searching an Excel workbook

```{r excel}
jtb_search_keyword(xlsx, "hives", columns = "comment")[
  , c("report_id", "site", "comment")]
```

Reading Excel needs the `readxl` package, which is a suggested dependency
rather than a hard one — the rest of `jtb2026` works without it, and asking for
an `.xlsx` without it produces an error that says exactly what to install.

### Choosing a worksheet

`sheet` takes a name or a 1-based position and is passed through to `readxl`.
It is ignored for every other input type, so it is harmless to leave in a
script that later switches to CSV.

```{r sheet, eval=FALSE}
jtb_search_keyword("study.xlsx", "headache", sheet = "Adverse events")
jtb_search_keyword("study.xlsx", "headache", sheet = 2)
```

Further arguments go to the reader too. That is how you skip a banner row above
the real header, a common shape for exported workbooks:

```{r skip, eval=FALSE}
jtb_search_keyword("study.xlsx", "headache", sheet = 1, skip = 3)

# The same idea for a delimited file, where the argument belongs to read.csv().
jtb_search_keyword("export.csv", "headache", skip = 3, na.strings = c("", "NA"))
```

### Column positions and file input

Positions refer to the columns **as read**, which is worth checking once when
the file has a banner row or a stray index column:

```{r positions}
names(jtb_read_table(csv))[c(5, 6)]

jtb_search_keyword(csv, "Nausea", columns = c(5, 6))$report_id
```

`jtb_read_table()` is exported precisely so you can look at what the search saw
before trusting a positional selection.

## Delimited files

```{r csv-in}
jtb_search_semantic(csv, "device malfunction",
                    columns = c("term", "comment"), top_n = 3)[
                      , c("report_id", "term", ".score")]
```

A CSV read this way goes through `utils::read.csv()`, so its type conversions
apply — `report_date` arrives as text rather than a `Date`, for instance. That
makes no difference to searching, which converts everything to text anyway, but
it matters if you plan to do arithmetic on the result.

```{r csv-types}
class(jtb_read_table(csv)$report_date)
class(site_reports$report_date)
```

## Writing the result out

`output_csv` writes the returned rows with `utils::write.csv(row.names =
FALSE)`. What you get in the file is what you get in R — including the score
columns from a semantic search, and the reporting columns from
`add_match_info = TRUE`.

```{r csv-out}
out <- file.path(tempdir(), "semantic_hits.csv")

hits <- jtb_search_semantic(xlsx, "stomach upset", columns = "comment",
                            top_n = 3, output_csv = out)

names(hits)
utils::read.csv(out)[, c("report_id", "comment", ".score", ".rank")]
```

The rows are written in the order they are returned, so a semantic result
arrives in the file already ranked — `.rank` 1 is the first data row.

If the target directory does not exist, the function stops **before** doing the
search rather than working and then failing at the last step:

```{r csv-out-error, error=TRUE}
jtb_search_keyword(site_reports, "headache",
                   output_csv = file.path(tempdir(), "nope", "hits.csv"))
```

```{r cleanup, include=FALSE}
unlink(out)
```

## Round trip

Nothing stops you writing the result of one search and searching it again — a
practical way to narrow a large export in two passes without holding the whole
thing in memory twice.

```{r round-trip}
step1 <- file.path(tempdir(), "boston.csv")

jtb_search_keyword(csv, "Boston", columns = "site", output_csv = step1)

jtb_search_semantic(step1, "head pain", columns = "comment", top_n = 2)[
  , c("report_id", "site", "comment", ".score")]
```

```{r round-trip-cleanup, include=FALSE}
unlink(step1)
```

## Errors worth recognising

```{r errors, error=TRUE}
jtb_search_keyword("no-such-file.csv", "headache")
```

```{r errors-2, error=TRUE}
jtb_search_keyword(site_reports, "headache", columns = "commnet")
```

```{r errors-3, error=TRUE}
jtb_search_keyword(site_reports, "headache", columns = 99)
```

Each one names the problem and, where it can, what the valid options were.
