---
title: "Get started with jtb2026"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    df_print: kable
vignette: >
  %\VignetteIndexEntry{Get started with jtb2026}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(width = 92)

# Render data frames as tables. `df_print: kable` above covers the vignette
# that R CMD build ships; pkgdown uses its own output format and ignores it,
# so register the method directly and both renders agree.
knit_print.data.frame <- function(x, ...) {
  knitr::asis_output(paste(c("", knitr::kable(x)), collapse = "\n"))
}
registerS3method("knit_print", "data.frame", knit_print.data.frame,
                 envir = asNamespace("knitr"))
```

> Find the rows of a table that mention what you are looking for.

`jtb2026` does one thing: you hand it a table and a word, and it hands back the
rows that mention it. The table can be an Excel workbook, a CSV, or a data
frame you already have; the answer can be returned to R, written to a CSV, or
both.

There are two ways to ask the question, and the difference between them is the
whole point of the package:

| | `jtb_search_keyword()` | `jtb_search_semantic()` |
| --- | --- | --- |
| Asks | *Does this row contain the word?* | *How close is this row to the idea?* |
| Returns | Every match, in the original row order | The best matches, ranked, with a score |
| Finds `"nauseous"` when you search `"stomach upset"` | No | Yes |
| Reproducible, offline, no model | Yes | Yes |

Each has a vignette of its own:
`vignette("keyword-search")`, `vignette("semantic-search")` and
`vignette("files-and-output")`.

## Install

```r
# install.packages("remotes")
remotes::install_github("yh202109/jtb2026", build_vignettes = TRUE)
```

```{r}
library(jtb2026)
```

## The mock table

The package ships `site_reports`, twenty-four invented rows of study-site
comments. Everything in these vignettes is computed against it. The `comment`
column is the interesting one — it is free text, written by different people
who described the same things in different words.

```{r mock-table}
str(site_reports)
```

```{r mock-head}
head(site_reports[, c("report_id", "site", "term", "comment")], 5)
```

## Sixty-second tour

### Find rows containing a word

```{r tour-keyword}
hits <- jtb_search_keyword(site_reports, "headache",
                           columns = c("term", "comment"))
hits[, c("report_id", "term", "comment")]
```

Four rows, in the order they appear in the table. Note the row names — `1`,
`5`, `14`, `21` — they are the positions in the input, so you can always trace
a result back.

### Find rows that *mean* something

Nobody in this table wrote "stomach upset". The keyword search says so:

```{r tour-keyword-miss}
nrow(jtb_search_keyword(site_reports, "stomach upset", columns = "comment"))
```

The semantic search finds all four nausea rows anyway, ranked:

```{r tour-semantic}
ranked <- jtb_search_semantic(site_reports, "stomach upset",
                              columns = "comment", top_n = 4)
ranked[, c("report_id", "term", "comment", ".score")]
```

`R009` mentions the stomach directly and scores highest. `R002` only says
"queasy" and scores lowest. The ordering is the useful part: it is a ranking,
not a yes/no.

### Write the answer to a CSV

Both functions take `output_csv`. The file gets exactly what the function
returns.

```{r tour-csv}
out <- file.path(tempdir(), "headache_rows.csv")

hits <- jtb_search_keyword(site_reports, "headache", output_csv = out)

nrow(hits)
readLines(out, n = 2)
```

```{r tour-csv-cleanup, include=FALSE}
unlink(out)
```

## Which columns get searched

Both functions take `columns`, and it accepts three shapes. All three of these
search the same two columns:

```{r columns}
by_name <- jtb_search_keyword(site_reports, "Nausea",
                              columns = c("term", "comment"))
by_position <- jtb_search_keyword(site_reports, "Nausea", columns = c(5, 6))

mask <- names(site_reports) %in% c("term", "comment")
by_mask <- jtb_search_keyword(site_reports, "Nausea", columns = mask)

identical(by_name, by_position)
identical(by_name, by_mask)
```

Positions are 1-based and refer to the columns of the table as read, so
`columns = c(5, 6)` is the fifth and sixth column. Leaving `columns` out
searches everything — convenient, but it means an identifier column can match
by accident:

```{r columns-all}
# "R002" is a report id, not a comment; searching every column finds it.
jtb_search_keyword(site_reports, "R002")[, c("report_id", "term")]
```

## Which one should you use?

Reach for `jtb_search_keyword()` when you know the exact string — a subject id,
a coded term, a site name — and you need *every* row that contains it. It is
exhaustive and its answer does not depend on any tuning.

Reach for `jtb_search_semantic()` when you are searching free text for an idea
and you do not know how the writer phrased it. It returns a ranked shortlist,
so treat it as triage: read the top of the list, not the whole thing.

A practical pattern is to use both. Rank first to find the vocabulary people
actually used, then keyword-search that vocabulary to get the complete set:

```{r combined}
shortlist <- jtb_search_semantic(site_reports, "stomach upset",
                                 columns = "comment", top_n = 5)
shortlist$comment

# Having read those, the words worth an exhaustive search are clear.
complete <- jtb_search_keyword(site_reports,
                               c("queasy", "nauseous", "stomach", "vomited"),
                               columns = "comment")
complete[, c("report_id", "comment")]
```

## Where to go next

- `vignette("keyword-search")` — whole words, case, regular expressions,
  several keywords at once, and knowing which column matched.
- `vignette("semantic-search")` — what the score actually measures, how to
  teach it your vocabulary, and how to plug in a real embedding model.
- `vignette("files-and-output")` — Excel sheets, delimited files, and what
  lands in the output CSV.
- `?jtb_search_keyword`, `?jtb_search_semantic` — every argument, in full.
