---
title: "Keyword search"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    df_print: kable
vignette: >
  %\VignetteIndexEntry{Keyword search}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(width = 92)

# Render data frames as tables. `df_print: kable` above covers the vignette
# that R CMD build ships; pkgdown uses its own output format and ignores it,
# so register the method directly and both renders agree.
knit_print.data.frame <- function(x, ...) {
  knitr::asis_output(paste(c("", knitr::kable(x)), collapse = "\n"))
}
registerS3method("knit_print", "data.frame", knit_print.data.frame,
                 envir = asNamespace("knitr"))
```

> Exhaustive, literal matching over the columns you choose.

```{r}
library(jtb2026)
```

`jtb_search_keyword()` returns every row whose selected columns contain the
term. Nothing is scored and nothing is reordered — the result is a subset of
the input, in the input's order, with the input's row names. That makes it the
right tool when you need completeness and traceability.

```{r basic}
hits <- jtb_search_keyword(site_reports, "headache", columns = "comment")
hits[, c("report_id", "comment")]
```

One row. Three other rows in this table are also headaches — their coded `term`
says so — but their comments say "head pain", "headche" and "Sharp pain across
the forehead". A literal search cannot know that any of those is the same
thing, and it is not pretending otherwise. That gap is what
`vignette("semantic-search")` is for.

## Case

Matching ignores case by default. Turn that off when the casing carries
meaning, such as a coded term or an identifier.

```{r case}
nrow(jtb_search_keyword(site_reports, "HEADACHE"))
nrow(jtb_search_keyword(site_reports, "HEADACHE", ignore_case = FALSE))
```

## Whole words

By default a keyword matches anywhere inside a value, so `"ache"` matches
`"headache"`. `whole_word = TRUE` requires word boundaries on both sides.

```{r whole-word}
jtb_search_keyword(site_reports, "ache", columns = "comment")$report_id

nrow(jtb_search_keyword(site_reports, "ache", columns = "comment",
                        whole_word = TRUE))
```

This matters most for short keywords and for identifiers that are prefixes of
one another.

## Keywords are literal, not patterns

Punctuation in a keyword is searched for, not interpreted. There is no need to
escape anything.

```{r literal}
doses <- data.frame(
  id = c("D1", "D2", "D3"),
  note = c("dose (2 mg) given", "dose 2 mg given", "wrote the c++ parser"),
  stringsAsFactors = FALSE
)

jtb_search_keyword(doses, "dose (2 mg)")
jtb_search_keyword(doses, "c++")
```

Set `regex = TRUE` when you do want a pattern. The keyword is then passed
through as a PCRE regular expression, and `whole_word` is ignored — write the
anchors yourself.

```{r regex}
jtb_search_keyword(doses, "dose [(]?[0-9]+ mg", regex = TRUE)

# Anything ending in a unit of mass.
jtb_search_keyword(doses, "\\d+ (mg|kg)\\b", regex = TRUE)$id
```

## Several keywords at once

Pass a character vector. `match = "any"`, the default, keeps rows matching at
least one keyword; `match = "all"` keeps rows matching every one.

```{r multiple-any}
jtb_search_keyword(site_reports, c("nausea", "fatigue"),
                   columns = "term")[, c("report_id", "term")]
```

Under `match = "all"` the keywords may land in *different* columns of the same
row, which is what makes it useful for narrowing:

```{r multiple-all}
jtb_search_keyword(site_reports, c("Boston", "Headache"),
                   match = "all")[, c("report_id", "site", "term")]
```

Asking for two keywords that never co-occur returns an empty table rather than
an error:

```{r multiple-all-empty}
nrow(jtb_search_keyword(site_reports, c("nausea", "fatigue"),
                        columns = "term", match = "all"))
```

## Knowing where the match was

With `add_match_info = TRUE` the result gains two reporting columns:
`.matched_columns` and `.matched_keywords`. Useful when you searched several
columns and want to know whether a hit came from the coded term or from someone's
free text.

```{r match-info}
jtb_search_keyword(site_reports, "headache", add_match_info = TRUE)[
  , c("report_id", "term", ".matched_columns")]
```

Only `R001` matched in both places. The other three matched the coded term
alone, because their comments say "head pain", "headche" and "Sharp pain across
the forehead" — three ways of writing a headache, none of them the word.

## Nothing found

A search with no matches returns a zero-row data frame with the original
columns. Downstream code can carry on without a special case.

```{r empty}
none <- jtb_search_keyword(site_reports, "unobtainium")
nrow(none)
names(none)
```

## What the call did

Every result carries a `jtb_search` attribute recording how it was produced.
Handy for logging, and for writing down in a report what was actually run.

```{r attribute}
str(attr(hits, "jtb_search"))
```

## Missing values

Empty cells are treated as empty text. An `NA` is never searchable as the
literal string `"NA"`, so a missing value cannot be mistaken for a match.

```{r missing}
partial <- data.frame(
  a = c("headache", NA),
  b = c(NA, "nausea"),
  stringsAsFactors = FALSE
)

nrow(jtb_search_keyword(partial, "headache"))
nrow(jtb_search_keyword(partial, "NA", ignore_case = FALSE))
```

## Writing the result out

`output_csv` writes exactly what is returned, including the reporting columns
when you asked for them. The directory has to exist; the function stops before
searching if it does not, rather than doing the work and then failing.

```{r csv}
out <- file.path(tempdir(), "keyword_hits.csv")

result <- jtb_search_keyword(site_reports, "headache",
                             columns = c("term", "comment"),
                             add_match_info = TRUE,
                             output_csv = out)

utils::read.csv(out)[, c("report_id", "term", ".matched_columns")]
```

```{r csv-cleanup, include=FALSE}
unlink(out)
```

See `vignette("files-and-output")` for reading Excel workbooks and delimited
files, and `?jtb_search_keyword` for the complete argument list.
