---
title: "Semantic search"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    df_print: kable
vignette: >
  %\VignetteIndexEntry{Semantic search}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(width = 92)
```

> Ranking rows by how close they are to an idea.

```{r}
library(jtb2026)
```

`jtb_search_semantic()` asks a different question from the keyword search: not
*does this row contain the word*, but *how close is this row to the idea*. It
returns rows sorted best-first with a `.score` column, so the result is a
shortlist to read rather than a set to trust wholesale.

```{r motivating}
jtb_search_semantic(site_reports, "stomach upset",
                    columns = "comment", top_n = 4)[
                      , c("report_id", "comment", ".score")]
```

None of those rows contains the phrase "stomach upset". Two do not even contain
the word "stomach".

## How the score is computed

There is no neural network here by default. The score is a blend of two
classical similarities, which is what keeps the function offline, fast and
deterministic — the same call gives the same answer next year.

### 1. Token similarity (75% of the score)

The query and every row are put through `jtb_tokenize()`: lower-cased, split on
punctuation, stripped of stop words, then stemmed so word endings stop
mattering.

```{r tokenize}
jtb_tokenize("Subject reported a severe HEADACHE after dosing.")
```

Compare that with the unstemmed version to see what the stemmer is doing:

```{r tokenize-nostem}
jtb_tokenize("Subject reported a severe HEADACHE after dosing.", stem = FALSE)
```

The tokens are then compared by **TF-IDF cosine similarity**. The IDF half is
what makes this better than counting shared words: a term that appears in
nearly every row carries almost no weight, while a rare term carries a lot. In
this table "subject" is nearly worthless and "autoinjector" is decisive.

Before the comparison, the query is expanded through a thesaurus. A query word
belonging to a synonym group brings in the rest of that group at
`synonym_weight` (0.6 by default, so a synonym counts for rather less than a
word you actually typed). That expansion is what carries "stomach" across to
"queasy":

```{r synonyms}
jtb_synonyms()$nausea
```

### 2. Fuzzy similarity (25% of the score)

The same cosine, computed over character 3-grams instead of words. This is the
half that survives a typo, an unusual plural or a word ending the stemmer did
not anticipate.

Row `R014` in the mock table misspells its own subject: *"Mild headche noted in
the diary"*. A correctly spelled query still reaches it.

```{r fuzzy}
fuzzy <- jtb_search_semantic(site_reports, "headache", columns = "comment",
                             top_n = NULL, min_score = 0.01)
fuzzy[, c("report_id", "comment", ".score")]
```

Turn the fuzzy half off and the misspelled row disappears:

```{r fuzzy-off}
jtb_search_semantic(site_reports, "headache", columns = "comment",
                    weights = c(token = 1, ngram = 0),
                    top_n = NULL, min_score = 0.01)$report_id
```

### Reading the score

`.score` runs from 0 to 1 and is **comparable within one call only**. IDF
depends on the table being searched, so a 0.4 here and a 0.4 against a
different table do not mean the same thing. Use the ranking, and use
`min_score` as a cut-off you tune by looking at results — not as an absolute
quality bar.

## Controlling how much comes back

`top_n` caps the number of rows; `min_score` drops weak ones. Both apply, and
`top_n = NULL` means "no cap".

```{r trimming}
# The default: at most the ten best rows, each scoring at least 0.05.
nrow(jtb_search_semantic(site_reports, "tired", columns = "comment"))

# Everything, scored, in rank order.
all_rows <- jtb_search_semantic(site_reports, "tired", columns = "comment",
                                top_n = NULL, min_score = 0)
nrow(all_rows)
head(all_rows$.score, 8)

# Only strong matches, however many that is.
jtb_search_semantic(site_reports, "tired", columns = "comment",
                    top_n = NULL, min_score = 0.2)[, c("report_id", "comment")]
```

`add_score = FALSE` leaves the input columns untouched if you want the rows
without the two extra columns.

## Teaching it your vocabulary

The built-in thesaurus is a small starter set. It will not know your domain,
and the honest failure mode is a query that returns nothing:

```{r miss}
nrow(jtb_search_semantic(site_reports, "skin irritation", columns = "comment"))
```

Neither "skin" nor "irritation" appears in the table, and neither is in a
default synonym group, so there is nothing to match on. Supply a group and the
same query works:

```{r custom-synonyms}
skin_terms <- list(
  skin = c("skin", "irritation", "irritated", "rash", "hives", "urticaria",
           "itchy", "itching", "patch", "redness", "swelling")
)

jtb_search_semantic(site_reports, "skin irritation", columns = "comment",
                    synonyms = skin_terms, top_n = 4)[
                      , c("report_id", "term", "comment", ".score")]
```

Your list is merged over the defaults **by group name**, so adding a group
keeps everything else working, and reusing a default name replaces that group:

```{r merge-semantics}
# The nausea group is still in force alongside the new skin group.
jtb_search_semantic(site_reports, "stomach upset", columns = "comment",
                    synonyms = skin_terms, top_n = 2)$report_id
```

Pass `synonyms = NULL` to switch expansion off entirely. It is worth doing once
on a real query, to see how much of the answer the thesaurus is responsible
for:

```{r no-synonyms}
with_thesaurus <- jtb_search_semantic(site_reports, "exhausted and sleepy",
                                      columns = "comment", top_n = 3)
without <- jtb_search_semantic(site_reports, "exhausted and sleepy",
                               columns = "comment", top_n = 3,
                               synonyms = NULL)

with_thesaurus[, c("report_id", "comment", ".score")]
without[, c("report_id", "comment", ".score")]
```

Only one row literally says "exhausted"; the thesaurus is what brings in the
row that says "unusually tired".

## Other knobs

```{r knobs, eval=FALSE}
jtb_search_semantic(
  site_reports, "device problem",
  columns = c("term", "comment"),
  weights = c(token = 0.9, ngram = 0.1),  # trust words more than spelling
  synonym_weight = 0.8,                   # synonyms nearly as good as typed words
  stopwords = character(),                # keep "no", "not", "off" ...
  stem = FALSE                            # exact word forms only
)
```

`stopwords = character()` is worth remembering. The default list contains
`"no"` and `"not"`, so out of the box *"no rash reported"* and *"rash
reported"* look alike. Pass an empty stop-word list when negation matters, and
the negations survive into the comparison.

## Using a real embedding model

For genuine model-based semantics, pass `embed_fun`: any function that takes a
character vector and returns a numeric matrix with one row per string. When it
is supplied the two similarities above are skipped and the score becomes the
cosine similarity between the query's embedding and each row's, rescaled to
0–1.

The shape of the contract is easiest to see with a toy embedding — one
dimension per letter:

```{r embed-toy}
letter_counts <- function(x) {
  t(vapply(strsplit(tolower(x), ""), function(ch) {
    tabulate(match(ch, letters), nbins = 26L)
  }, numeric(26)))
}

jtb_search_semantic(site_reports, "headache", columns = "comment",
                    embed_fun = letter_counts, top_n = 3)[
                      , c("report_id", "comment", ".score")]
```

Those results are nonsense, which is the point: the function does exactly what
your embedding tells it to, and a bad embedding gives bad rankings. A real one
looks the same from the package's side:

```{r embed-real, eval=FALSE}
# Sketch: replace with whichever embedding service or local model you use.
embed <- function(x) {
  vecs <- lapply(x, function(txt) my_embedding_api(txt))  # numeric vectors
  do.call(rbind, vecs)
}

jtb_search_semantic("reports.xlsx", "device malfunction",
                    columns = "comment", embed_fun = embed, top_n = 20)
```

Two things to keep in mind. The function is called **once** with the query
followed by every row, so an API-backed embedding should batch internally and
you should cache it if you search the same table repeatedly. And the result
stops being reproducible: it is now only as stable as the model behind it.

## What it will not do

- It does not understand negation, sarcasm or quantity. "No headache" and
  "severe headache" score similarly.
- It has no idea what your abbreviations mean unless you put them in
  `synonyms`.
- Ranking is not retrieval. If you need *every* row that mentions a term —
  for a count, an audit, a regulatory listing — use `jtb_search_keyword()`,
  which is exhaustive by construction.

A good working pattern is to rank first to learn the vocabulary, then
keyword-search that vocabulary for the complete set. `vignette("jtb2026")`
shows it end to end.
